---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
#beolvassuk a csv-t
expeditions <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv')
#megpróbáljuk rendezni, köszi a forcats tippet
top_peaks <- expeditions %>%
  #megtartjuk a top 15 csúcsot, elvetjük az other-t
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other") %>%
  #kezdjük előkészíteni a rangsort
  count(peak_name, season, name = "n_expeditions") %>%
  group_by(peak_name) %>%
  mutate(total_expeditions = sum(n_expeditions)) %>%
  #ungroup nélkül nem lesz jó a sorrend
  ungroup() %>%
  mutate(
    peak_name = fct_reorder(peak_name, total_expeditions),
    season = fct_relevel(season, c("Autumn", "Spring", "Summer","Unknown", "Winter")))
#elkezdjük a plotot
ggplot(top_peaks, aes(x = n_expeditions, y = peak_name, fill = season)) +
  geom_col() +
  #hozzáadjuk a veridris témát
  scale_fill_viridis_d(name = "season") +
  #megadjuk a címeket és a pozíciókat
  labs(
    x = NULL,
    y = "Number of expeditions",
    title = "The 15 most popular peaks stacked by season of expedition") +
    #kiszedjük a szürke hátteret a táblából
  theme_light() +
  #jobbról az aljára tesszük a color coding-ot  
  theme(legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}

phd_by_field <- tidytuesdayR::tt_load("2019-02-19")$phd_by_field
#phd_by_field tidynak tűnik, csoportosítunk a releváns változók alpján
phd_all <- phd_by_field %>%
group_by(year, broad_field) %>%
#összegezünk és figyelmen kívül hagyjuk a NA értékeket
summarise(phds = sum(n_phds, na.rm = TRUE))
#plotot készítjük a megadott paraméterek alapján
ggplot(phd_all, aes(x = year, y = phds, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_color_brewer(palette = "Dark2", name = "Broad field") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(
    title = "Number of awarded Ph.D.-s in the US by year",
    y = NULL,
    x = NULL

  ) +
  theme_minimal()
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
# GitHub API rate limit exceeded, fapadosan olvassuk be
commute <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-11-05/commute.csv")

state_commutes <- commute %>% 
  #ide csak azért van betéve a state_region hogy tudjak majd színként hivatkozni rá, a state_abb is a plot miatt van itt
  group_by(state, state_region, state_abb) %>%
  summarise(
    BIKE = sum(n[mode == "Bike"], na.rm = TRUE),
    WALK = sum(n[mode == "Walk"], na.rm = TRUE)
  )

ggplot(state_commutes, aes(x = WALK, y = BIKE, color = state_region)) +
  geom_point() + 
  geom_text(aes(label = state_abb),color = "black")+
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(
    title = "Title number of people walking vs. biking to work in each USA state",
    x = "Number of ppl walking to work (log N)",
    y = "Number of ppl biking to work (log N)"
  ) +
    theme_minimal()
#hát van itt néhány állam pluszban, de aes és geom szempontból szerintem hozza az elvártat

```
